---
title: "IATI Data Analysis Demonstration Notebook"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 3
params:
  iati_file: "data/sample_iati.xml"
  group_by: "recipient-country"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(xml2)
library(dplyr)
library(ggplot2)
library(tibble)
library(stringr)
library(purrr)
```

<style>

/* ----------------------------- */
/* PROFESSIONAL, BALANCED LAYOUT */
/* ----------------------------- */

body {
  font-size: 15px;
  line-height: 1.45;
  font-family: "Helvetica Neue", Arial, sans-serif;
  color: #222;
}

.main-container {
  max-width: 900px;
  margin: auto;
  padding: 20px 40px;
}

/* Headings: clean, consistent, not oversized */
h1 {
  font-size: 28px;
  margin-top: 35px;
  margin-bottom: 15px;
  font-weight: 600;
}
h2 {
  font-size: 22px;
  margin-top: 28px;
  margin-bottom: 10px;
  font-weight: 600;
}
h3 {
  font-size: 18px;
  margin-top: 22px;
  margin-bottom: 8px;
  font-weight: 600;
}

/* Paragraph spacing */
p {
  margin: 8px 0 12px 0;
}

/* Tables */
table {
  font-size: 14px;
}

/* Code blocks */
pre, code {
  font-size: 13px;
}

/* Expandable code sections */
.ods-details > summary {
  cursor: pointer;
  font-weight: 600;
  margin-bottom: 6px;
}
.ods-details {
  margin-bottom: 18px;
}

/* Hover tooltip with icon */
.hover-term {
  border-bottom: 1px dotted #1a73e8;
  cursor: help;
  color: #1a73e8;
  font-weight: 600;
  padding: 0 2px;
}
.hover-term:hover {
  background-color: #eef5ff;
}
.hover-icon {
  font-weight: bold;
  margin-right: 3px;
  color: #1a73e8;
}

.authorship-note {
  font-size: 14px !important;
  font-style: italic !important;
  color: #555 !important;
  line-height: 1.4;
  margin: 10px 0 20px 0;
}



</style>

# Purpose of This Notebook

This notebook has been created as part of my application to **Open Data Services (ODS)**.  
Its purpose is to demonstrate:

- ability to design and structure a clear, reproducible analytical workflow  
- understanding of notebook‑based analysis as a transparent communication tool  
- approach to validation, documentation, and open‑data‑aligned practices  
- ability to collaborate with tools (including AI) while maintaining full analytical oversight  

> **Authorship note:**  
<div class="authorship-note">
This notebook was developed through a collaborative workflow combining human expertise and AI‑assisted drafting.  
All analytical decisions, validation, and final editorial judgement were carried out by **Gillian Paterson**.  
AI assistance supported idea generation, code structuring, and iterative refinement.
</div>


---

# Overview

This notebook provides a **pseudo‑analysis** of synthetic 
<span class="hover-term" 
      title="The International Aid Transparency Initiative (IATI) is a global open‑data standard for publishing information about development and humanitarian activities. It provides a structured, machine‑readable way for organisations to share data on funding, projects, locations, and results.">
<span class="hover-icon">ℹ️</span>IATI
</span> data, designed to demonstrate an ODS‑style approach to clear, transparent, and reproducible analytical workflows.


> **Note:** This report uses *synthetic* IATI data for demonstration purposes only.

---

# 1. Load and Parse IATI XML

<details class="ods-details">
<summary><strong>Show code</strong></summary>

```{r, echo=TRUE}
doc <- read_xml(params$iati_file)

activities <- xml_find_all(doc, ".//iati-activity")

get_text <- function(node, xpath) {
  x <- xml_find_first(node, xpath)
  if (is.na(x)) NA_character_ else xml_text(x)
}

df_activities <- tibble(
  activity_id        = sapply(activities, get_text, ".//iati-identifier"),
  reporting_org      = sapply(activities, get_text, ".//reporting-org"),
  recipient_country  = sapply(activities, get_text, ".//recipient-country/@code"),
  sector             = sapply(activities, get_text, ".//sector/@code"),
  title              = sapply(activities, get_text, ".//title/narrative")
) |> distinct()
```

</details>

### Extracted activities

```{r}
df_activities
```

---

# 2. Transactions

<details class="ods-details">
<summary><strong>Show code</strong></summary>

```{r, echo=TRUE}
extract_transactions <- function(act_node) {
  act_id <- get_text(act_node, ".//iati-identifier")
  txs <- xml_find_all(act_node, ".//transaction")
  if (length(txs) == 0) return(NULL)
  
  tibble(
    activity_id = act_id,
    tx_type     = sapply(txs, get_text, ".//transaction-type/@code"),
    tx_value    = as.numeric(sapply(txs, get_text, ".//value")),
    tx_currency = sapply(txs, get_text, ".//value/@currency"),
    tx_date     = sapply(txs, get_text, ".//transaction-date/@iso-date")
  )
}

df_tx <- map_dfr(activities, extract_transactions)
```

</details>

### Extracted transactions

```{r}
df_tx
```

---

# 3. Validation Checks

<details class="ods-details">
<summary><strong>Show code</strong></summary>

```{r, echo=TRUE}
valid_countries <- c("KE", "UG", "RW", "TZ")

validation_summary <- tibble(
  check = c(
    "Missing activity_id",
    "Missing reporting_org",
    "Missing recipient_country",
    "Missing sector",
    "Missing title",
    "Duplicate activity_id",
    "Invalid recipient_country code",
    "Activities with no transactions"
  ),
  count = c(
    sum(is.na(df_activities$activity_id) | df_activities$activity_id == ""),
    sum(is.na(df_activities$reporting_org) | df_activities$reporting_org == ""),
    sum(is.na(df_activities$recipient_country) | df_activities$recipient_country == ""),
    sum(is.na(df_activities$sector) | df_activities$sector == ""),
    sum(is.na(df_activities$title) | df_activities$title == ""),
    sum(duplicated(df_activities$activity_id)),
    sum(!df_activities$recipient_country %in% valid_countries),
    sum(!df_activities$activity_id %in% df_tx$activity_id)
  )
)
```

</details>

### Validation summary

```{r}
validation_summary
```

---

# 4. Grouped Activity Summary

Grouping by:

**`r params$group_by`**

<details class="ods-details">
<summary><strong>Show code</strong></summary>

```{r, echo=TRUE}
group_var <- params$group_by
group_var_clean <- gsub("-", "_", group_var)

grouped_activities <- df_activities |>
  group_by(.data[[group_var_clean]]) |>
  summarise(
    n_activities = n(),
    .groups = "drop"
  ) |>
  arrange(desc(n_activities))
```

</details>

### Summary table

```{r}
grouped_activities
```

---

# 5. Charts

## Activities by group

```{r}
ggplot(grouped_activities, aes(
  x = reorder(.data[[group_var_clean]], n_activities),
  y = n_activities
)) +
  geom_col(fill = "#005ea5") +
  coord_flip() +
  theme_minimal() +
  labs(
    x = params$group_by,
    y = "Number of activities",
    title = paste("Activities grouped by", params$group_by)
  )
```

## Total transaction value by country

```{r}
df_tx_joined <- df_tx |>
  left_join(df_activities, by = "activity_id")

tx_by_country <- df_tx_joined |>
  group_by(recipient_country) |>
  summarise(
    total_value = sum(tx_value, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(tx_by_country, aes(
  x = reorder(recipient_country, total_value),
  y = total_value
)) +
  geom_col(fill = "#28a197") +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Country",
    y = "Total value",
    title = "Total transaction value by country"
  )
```

